---
title: "Juvenile Snowshoe Hares in the Bonanza Creek Experimental Forest"
author: "Taylor Medina"
date: "11/2/2021"
output: 
  html_document: 
    toc: yes
    theme: paper
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(here)
library(kableExtra)
library(lubridate)
library(plotly)
library(effsize)
library(broom)
library(equatiomatic)
```

# Introduction
This report provides an exploratory analysis of juvenile snowshoe hares in the Bonanza Creek experimental forest in Alaska  observed from 1999 - 2012, provided by the work of Dr. Knut Kielland (Kielland et. al. 2017.). A statistical analysis is provided of the number of juvenile hare trappings performed for each year of the study, as well as analysis on the relationship between site and gender on the weight of a hare, as well as a linear regression analysis on the relationship between the mass of the hare and its hind foot length. 

# Data and Analyses
The Bonanza Hares data set provided by Dr. Kielland (Kielland et. al. 2017) contains data on hare trappings in three different sites in the Bonanza Creek: Riparian forest, Mature Forest, and Lowlands. The hare trapping information contains data on the time and date of the trapping, the trap-specific id, as well as the sex, age, ear length, and the hind-foot length of the hare. The trappings took place from 1999 - 2012. For the exploratory data visualization, mean masses between female and male juvenile hares were compared using a standard 2-sample unpaired t-test with an $\alpha$ = 0.05. The relationship between weight and hind foot length was explored using a simple ordinary least squares regression. All analyses are on R Version 4.0.2. using RStudio Version 1.4.1717.

# Juvenile Hare Investigation
## Annual Juvenile Hare Trap Counts
```{r, message = FALSE}
bonanza_hares <- read_csv(here("data", "bonanza_hares.csv")) %>% 
  filter(age == "j")

hare_trappings <- bonanza_hares %>% 
  mutate(date = mdy(date), # convert class of date column
         year = year(date)) %>% # splice out year
  group_by(year) %>% 
  summarize(trappings = n()) # get number of trappings per year
```

```{r}
hare_plot <- ggplot(data = hare_trappings,
                    aes(x = year, y = trappings)) +
  geom_point() +
  geom_line() +
  theme_classic() +
  labs(x = "year",
       y = "number of trappings")

ggplotly(hare_plot)
```
**Figure 1:** Time series plot of the total number of hare-trappings performed from 1998 - 2012 (Data: Kielland et al. 2017)

```{r}
hare_summary <- hare_trappings %>% 
  summarize(mean_trap = round(mean(trappings), 0),
            sd_trap = round(sd(trappings), 0),
            median_trap = median(trappings),
            max_trap = max(trappings),
            min_trap = min(trappings))
```

The maximum number of trappings per year is `r hare_summary$max_trap` which was done in 2009. The minimum number of trappings per year is `r hare_summary$min_trap` which was done in 2002. The average number of trappings done per year over the experiment is `r hare_summary$mean_trap` and the median number of trappings is `r hare_summary$median_trap`. There are random amount of trappings performed per year and there are no years with the same amount of trappings. This could be due to the difficulty of trapping young hares or seasonal variability in their populations. To standardize the experiment a bit better, the same number of traps and the same time-frames for the trappings should be done every year in order to account for population variance over different seasons and decrease the likelihood of having such a high sampling range depending on the year,


## Visualize Juvenile Hare Weights
```{r, message = FALSE, warning = FALSE}
hare_weights <- bonanza_hares %>% 
  group_by(grid, sex) %>% 
  summarize(mean_weight = round(mean(weight, na.rm = TRUE),1),
            sd_weight = round(sd(weight, na.rm = TRUE),1)) %>%
  mutate(grid = case_when( # replace short names of sites with full names
    grid == "bonbs" ~ "Bonanza Lowland",
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Bonanza Riparian"))
 
x_axis <- c("female", "male", "NA") # define x-axis labels since they are strings

weight_plot <- ggplot(data = hare_weights,
                      aes(x = sex, y = mean_weight)) +
  geom_bar(stat = "identity", aes(fill = sex)) +
  geom_errorbar(data = hare_weights, 
                mapping = aes(x = sex, ymin = mean_weight - sd_weight, 
                              ymax = mean_weight + sd_weight, size = 0.2, width = 0.1)) +
  facet_wrap(~ grid) +
  theme_classic() +
  scale_y_continuous(expand = c(0,1)) +
  scale_x_discrete(labels = x_axis) + # manually define x-axis
  theme(legend.position = "none") +
  labs(x = "sex",
       y = "mass (g)")
  

ggplotly(weight_plot)
```
**Figure 2:** Bar chart comparing the average masses of female (red), male (blue), and unknown sex juvenile hares (grey). The error bars represent the standard deviation of the sample masses. NA represents an unknown gender. (Data: Kielland et al. 2017)

Overall, according to **Figure 2**, male juvenile hares have the highest weight compared to females and unknown sex juvenile hares. In the Bonanza Lowland site, females are lighter than males with a mass of `r hare_weights$mean_weight[1]` $\pm$  `r hare_weights$sd_weight[1]`g and `r hare_weights$mean_weight[2]` $\pm$  `r hare_weights$sd_weight[2]` g respectively. There was not enough data in this site to get a mean for unknown sex hares. There was only one sample collected and there was no mass measurement. In the Bonanza Mature site, male hares have a mass of `r hare_weights$mean_weight[5]` $\pm$  `r hare_weights$sd_weight[5]` g which is higher than females with a mass of `r hare_weights$mean_weight[4]` $\pm$  `r hare_weights$sd_weight[4]`g and unknown sex hares with a mass of `r hare_weights$mean_weight[6]` $\pm$  `r hare_weights$sd_weight[6]` g. In the Bonanza Riparian site, males are bigger with a mass of `r hare_weights$mean_weight[8]` $\pm$  `r hare_weights$sd_weight[8]` compared to females with a mass of `r hare_weights$mean_weight[7]` $\pm$  `r hare_weights$sd_weight[7]`g and unknown sex hares with a mass of `r hare_weights$mean_weight[9]` $\pm$  `r hare_weights$sd_weight[9]` g.

## Juvenile Weight Comparison
```{r}
hare_stats <- bonanza_hares %>% 
  group_by(sex) %>% 
  summarize(mean_weight = round(mean(weight, na.rm = TRUE), 1),
            sd_weight = round(sd(weight, na.rm = TRUE), 2),
            sample_size = n()) %>% 
  filter(sex != "") %>% # remove na row
  mutate(sex = case_when(
    sex == "f" ~ "female",
    sex == "m" ~ "male"))

  
hare_stats %>%
  kable(col.names = c("sex",
                      "mean weight (g)",
                      "standard deviation (g)",
                      "sample size"),
        caption = "Table 1: Mean weights with standard deviation of Bonanza Hares across all sampling sites (Data: Kielland et al. 2017)") %>% 
  kable_classic(full_width = FALSE) %>% 
  row_spec(0:2, align = "center")
```

```{r, include = FALSE}
hare_stats_1 <- bonanza_hares %>% 
  filter(age == "j") %>% 
  group_by(sex) %>% 
  summarize(weight = weight) 

ggplot(data = hare_stats_1, aes(x = weight)) +
  geom_bar() + 
  facet_wrap(~ sex)

ggplot(data = hare_stats_1, aes(sample = weight)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ sex)
```

```{r}
female_hares <- subset(hare_stats_1, sex == "f") %>% 
  na.omit() # omit na values

male_hares <- subset(hare_stats_1, sex == "m") %>% 
  na.omit() # omit na values

mean_diff = hare_stats$mean_weight[2] - hare_stats$mean_weight[1] # calculates difference in mean female vs. male height
cohen <- cohen.d(female_hares$weight, male_hares$weight) # performs a cohen D test to get the effect size
ttest <- t.test(female_hares$weight, male_hares$weight) # performs an unpaired 2-sample t-test

hare_analysis <- as.data.frame(t(c(mean_diff, cohen$estimate, ttest$p.value))) # puts results into a data frame
```
The average weight for juvenile female hares across all three sites is `r hare_stats$mean_weight[1]` $\pm$ `r hare_stats$sd_weight[1]` and for males is `r hare_stats$mean_weight[2]` $\pm$ `r hare_stats$sd_weight[2]` (see Table 1).
Male juvenile hares have a slightly higher mean mass than female hares according to **Table 1**, with the difference in these mean masses being `r round(mean_diff,1)` grams. To determine if this difference was statistically significantly, a standard unpaired two sample t-test was performed. The null hypothesis was that the difference in means between males and females is 0. The alternate hypothesis is that the difference in mean masses is not 0. The t-test was performed and a p-value of `r round(ttest$p.value, 3)` (effect size = `r abs(round(cohen$estimate,4))`) was found, meaning there is a `r round(ttest$p.value * 100, 3)`% chance that two samples of different gender juvenile hares would have a difference in means of at least `r round(mean_diff, 1)`, assuming the two samples are from the sample population. With such a low p-value, the null hypothesis is rejected since it is below the selected confidence interval ($\alpha$ = 0.05), meaning that the male population is heavier than the female population.

## Relationship between juvenile weight & hind foot length
```{r, warning = FALSE}
hare_foot <- bonanza_hares %>% 
  select(sex, weight, hindft, age) %>% 
  filter(age == "j") %>% 
  na.omit()

foot_plot <- ggplot(data = hare_foot, aes(x = weight, y = hindft)) +
  geom_point() +
  theme_classic() +
  labs(x = "weight (g)",
       y = "hind foot length (mm)") 
                      
ggplotly(foot_plot)                
```
**Figure 3:** Scatterplot showing the relationship between hare weight in grams compared to their hind foot length in millimeters.(Data: Kielland et al. 2017)

```{r, message = FALSE, warning = FALSE}
lm_plot <- ggplot(data = hare_foot, aes(x = weight, y = hindft)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic() +
  labs(x = "weight (g)",
       y = "hind foot length (mm)")

ggplotly(lm_plot)
```
**Figure 4:** Ordinary Least Squares Regression showing the relationship between hare weight in grams and hind foot length in millimeters. (Data: Kielland et al. 2017)

```{r, warning = FALSE, message = FALSE}
hare_lm <- lm(hindft ~ weight, data = hare_foot) # performs ordinary least squares regression

lm_tidy <- tidy(hare_lm)

lm_glance <- glance(hare_lm) %>% 
  mutate(beta_0 = lm_tidy$estimate[1], # extracts intercept value from model fit
         beta_1 = lm_tidy$estimate[2], # extracts slope value from model fit
         pearson = cor(hare_foot[3], hare_foot[2], method = "pearson")) %>% # calculates Pearson Correlation
  select(beta_1, beta_0, adj.r.squared, pearson)

lm_glance %>%
  round(4) %>% 
  kable(col.names = c("slope",
                      "intercept",
                      "$R^2$",
                      "Pearson correlation"),
        caption = "Table 2: Linear Regression Summary (Data: Kielland et al. 2017) ") %>% 
  kable_classic(full_width = FALSE) %>% 
  row_spec(0:1, align = "center")
```

```{r, include = FALSE}
plot(hare_lm)
```
The linear model that displays the relationship between the weight of a hare and its hind foot length is shown below:
```{r}
extract_eq(model = hare_lm, use_coefs = TRUE)
```

A slope ($\beta_1$) of `r lm_glance$slope` with units of $\frac{mm}{g}$ shows that for each unit increase in weight, the hind foot length of the hare will increase by `r lm_glance$slope` mm. An intercept ($\beta_0$) of `r lm_glance$intercept` shows that, though impossible, a hare with a mass of 0 would have a hind foot length of `r lm_glance$intercept` mm. The outcome of the linear regression displayed a weak linear relationship between the weight of a hare and its' hind foot length. Only `r lm_glance$adj.r.squared * 100`% of the variation in the scatter plot could be explained by the model according to **Table 4**. The Pearson Correlation of `r lm_glance$pearson` shows that there is a moderate positive correlation between weight and hind foot length. Assumptions of homoscedasticity is of concern in this model as the residuals are not relatively constant in the model. Additionally, the assumption that the distribution of errors is normally distributed is of concern here. Furthermore, based off of the scatter plot, the data does not seem to follow a linear relationship. 

# Conclusion
From the analysis of the hare weights by gender, there was a large variation in the average weights of each sex in each sampling site, but overall the male juvenile hares have the highest weight. The average weight (in grams) of the captured juvenile male hares is `r hare_stats$mean_weight[2]` and for females is `r hare_stats$mean_weight[1]` with the mean difference being `r mean_diff`. From performing a two sample unpaired t-test,  p-value of `r round(ttest$p.value, 3)` (effect size = `r abs(round(cohen$estimate,4))`) was found, which is smaller than the confidence level of 0.05, rejecting the null hypothesis that the difference in mean masses of males and females is 0, meaning that males are heavier than females overall. 
The regression analysis of the relationship between the weight of a hare and its hind foot length showed a weak positive linear correlation with the linear model having an $R^2$ = `r round(lm_glance$adj.r.squared,3)`, which means almost 75% of the variance is not accounted for by the linear model. The final conclusion for this analysis is that there is not a linear relationship between hare weight and foot length, meaning there are other factors that influence this length or there is another model fit that would fit better.

# References
Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative